---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

#The P(e):P(o) package
This is the alpha release of `pepo`, a minimal package to calculate the probabilities of hemiplasy (trait evolution incongruent with species tree due to incomplete lineage sorting) and homoplasy (incongruent traits due to convergent mutations). Currently, the package includes functions to estimate the ratio of these two probabilities, which we call the Hemiplasy Risk Factor (HRF), for all branches on a phylogeny. Provided a tree, branch lengths and population-wide mutation rate in coalescent units (2N), the HRF will give an intuition for the relative importance of hemiplasy in the evolution of a particular clade.

## Installation
This package depends on `ape`, `dplyr`, and `purrr`. The last two are part of the `tidyverse`. 

Install `pepo` from github with: `devtools::install_github("guerreror/pepo")`.

```{r, warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)
library(pepo)
```

## Quick example

In this example we start with a preloaded phylogeny of Solanum sect Lycopersicon from Pease et al (2016). The tree is already of class `phylo` (from the 'ape' package). 
=======
## Minimal example

In this example, we start with a preloaded phylogeny of Solanum sect Lycopersicon from Pease et al (2016). The tree is already of class `phylo` (from the 'ape' package). 

```{r, echo=TRUE}
data("tomato")
class(tomato)
```

The two functions you'll need from `pepo` are: `prep_branch_lengths()` and `tree_hrf()`. The former returns a tibble (a tidy data frame) with variables that will be needed by the latter.

```{r, echo = TRUE, warning=FALSE, message=FALSE}
tomato_branches <- prep_branch_lengths(tomato) 
tomato_branches
```

Then we can call `tree_hrf()` on that tibble. The function will return the original data frame plus a new variable, `hrf`. This function assumes branch lengths are **in coalescent units** (e.g., calculated in MP-EST). The call below will assume the default population-wide mutation rate (0.01).

```{r, echo = TRUE}
tomato_hrf <- tree_hrf(tomato_branches)
tomato_hrf
```

Some `NA` values in the `hrf` column are normal: the function does not calculate HRF for tips or ancestral branches. This is because the HRF is a property of a branch that has: 1) two descendant lineages, 2) a sister lineage, and 3) an ancestral branch with known length.

That's it. Now we can explore/plot the HRF of all branches in the phylogeny. For example, we can use the `ggtree` package to plot the tree. The `to_treedata()` function converts our HRF tibble and `phylo` tree into a `ggtree`-compatible object (which allows for easy plotting).

```{r, error=FALSE, message=F, warning=F}
library(ggtree)
solgg <- to_treedata(tomato, tomato_hrf%>%
  mutate(cathrf = cut(hrf, breaks=c(0, 0.2, 0.5, 0.8, 1))))
library(rcartocolor)
solplot <- ggtree(solgg, aes(color=hrf), size=2) + 
  geom_tiplab(color='black') +
  scale_color_gradient2(limits=c(0,1), low='#008080', mid='#f6edbd', high='#ca562c', midpoint=0.5, na.value = 'grey90')+
  theme(legend.position = c(.05, .85))

ggsave(solplot, filename = 'solplot.pdf', height=8, width=5)

```

##Hemiplasy Risk in the Great Apes

This is a quick calculation of the HRF on the phylogeny of Human-Chimpanzee-Gorilla.

```{r, echo=TRUE}
apetree <- ape::read.tree(text = "((((human,chimpanzee),gorilla),orangutan),out);")
#nodevec <- c("human","chimpanzee", "gorilla", "orangutan", "hcg", "hc")
#two_n_vec <- c(125, 45, 65, 10, 20, 21, 19, 50)*1000*2 #2Ne for each node
two_n_vec <- c(125, 45, 65, 15, 15, 18, 19, 50)*1000*2
pop_std_vec <- two_n_vec/two_n_vec[3] #relative to human 2Ne
time_vec <- c(5, 6.5, 1.5, 4.1, 4.1, 5.5, 12, 15)*1000000 # times in MY
length_vec <- time_vec/(20 * two_n_vec) # coalescent branch lengths: MY/(gen * 2Ne)
apetree$edge.length <- length_vec
plot(apetree); ape::edgelabels(prettyNum(length_vec, digits=2))

```

```{r, echo = TRUE}
ape_hrf <- prep_branch_lengths(apetree)%>% tree_hrf(mutation =(2*10000*10^-8))
ape_hrf
```

```{r, error=FALSE, message=F, warning=F}
hc_hrf <- ape_hrf %>% filter(to==9)

# the data here are from Kong et al 2012
# numerators are sums of CpG and nonCpG mutations, from Table 2
# the denominator 2*78*2.583e9 is the number of sites observed (effective bases times transmission events -- two times number of trios)

cpg <- tibble(mu =c(1.2e-8, (2489 + 1516)/(2*78*2.583e9), (855+ 73)/(2*78*48.8e6)))%>%
  mutate(nmu = 2*15000*mu)%>%
  mutate(hc =  map_dbl(nmu, function(x) tree_hrf(hc_hrf, mutation = x, mode = 'strict', pepo=T)$hrf))
cpg
```

