---
title: "Hemiplasy Risk Factor for a phylogeny"
author: "Rafael F. Guerrero"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
#The P(e):P(o) package
This is the alpha release of `pepo`, an R package to calculate the hemiplasy risk factor (HRF) on a phylogeny. This packages depends on `ape`, `dplyr`, and `purrr`. The last two are part of the `tidyverse` package (loaded below). 

```{r, warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)
library(pepo)
```

In this example, we start with a preloaded phylogeny of Solanum sect Lycopersicon from Pease et al (2016). The tree is already of class `phylo` (from the 'ape' package). 

```{r, echo=TRUE}
data("tomato")
class(tomato)
```

## Minimal example
The two functions you'll need from `pepo` are: `prep_branch_lengths()` and `tree_hrf()`. The former returns a tibble (or `data_frame`) with variables that will be needed by the latter.

```{r, echo = TRUE}
tomato_branches <- prep_branch_lengths(tomato) 
tomato_branches
```

Then we can call `tree_hrf` on that data frame. The function will return the original data frame plus a new variable, `hrf`. This function assumes branch lengths are **in coalescent units** (e.g., calculated in MP-EST). The call below will assume the default population-wide mutation rate (0.01).

```{r, echo = TRUE}
tomato_hrf <- tree_hrf(tomato_branches)
tomato_hrf
```

That's it. Now we can explore/plot the HRF of all branches in the phylogeny. For example, we can use the `ggtree` package to plot the tree. `pepo` includes the `to_treedata` function to convert our data frame and `phylo` tree into a `ggtree`-compatible object.

```{r, error=FALSE, message=F, warning=F}
library(ggtree, quietly=T)
solgg <- to_treedata(tomato, tomato_hrf)

ggtree(solgg, aes(color=hrf), size=2)+geom_tiplab(color='black')+
  scale_color_gradient2(low = "darkblue", mid = "yellow", high="red", midpoint=1, na.value="grey80")+
  theme(legend.position = c(.05, .85))
```
